---
title: "Predicting Race Using Demographic Information"
output:
  pdf_document: default
  html_document: default
---

In this exercise, we return to the task of predicting the ethnicity of individual voters given their surname and residence location. We'll be using Bayes' rule to draw these inferences. This exercise is based on the following article: Kosuke Imai and Kabir Khanna. (2016). ["Improving Ecological Inference by Predicting Individual Ethnicity from Voter Registration Records."](https://doi.org/10.1093/pan/mpw001) *Political Analysis* 24(2): 263-272.

In this exercise, we attempt to improve our prediction by taking into account demographic information such as age and gender. As done earlier, we validate our method by comparing our predictions with the actual race of each voter.

------------------------------------------------------------------------------
 Name                Description
 ------------------- ---------------------------------------------------------
 `county`            County census id of voting district.
 
 `VTD`               Voting district census id (only unique within county)
 
 `total.pop`         Total population of voting district
------------------------------------------------------------------------------

Other variables are labeled in three parts, each separated by a period. See below for each part. Each column contains the proportion of people of that gender, age group, and race in the voting district. 

------------------------------------------------------------------------------
 Name                Description
 ------------------- ---------------------------------------------------------
 `gender`            Male or female
 
 `age groups`        Age groups as defined by U.S. Census (see table below)
 
 `race`              Different racial categories (see table below)
------------------------------------------------------------------------------

 Below is the table for variables describing racial categories:

------------------------------------------------------------------------------ 
  Name               Description
 ------------------- ---------------------------------------------------------
 `whi`               non-Hispanic whites in the voting district
 
 `bla`               non-Hispanic blacks in the district
 
 `his`               Hispanics
 
 `asi`               non-Hispanic Asian and Pacific Islanders
 
 `oth`               other racial categories
 
 `mix`               non-Hispanic people of two or more races.
------------------------------------------------------------------------------
 
 Below is the table for age-group variables, as defined by the U.S. Census: 
 
------------------------------------------------------------------------------
 Name                Description
 ------------------- ---------------------------------------------------------
 `1`                 18--19
 
 `2`                 20--24
 
 `3`                 25--29
 
 `4`                 30--34
 
 `5`                 35--39
 
 `6`                 40--44
 
 `7`                 45--49
 
 `8`                 50--54
 
 `9`                 55--59
 
 `10`                60--64
 
 `11`                65--69
 
 `12`                70--74
 
 `13`                75--79
 
 `14`                80--84
 
 `15`                85+
------------------------------------------------------------------------------
 
We use three data sets in this exercises, two of which were already introduced in Section 6.1. 
The first data set is a random sample of 10,000 registered voters contained in the csv file, `FLVoters.csv`. *Table 6.1* presents the names and descriptions of variables for this data set.
The second data set is a csv file, `cnames.csv`, containing a modified version of the original data set, `names.csv`, after making appropriate adjustments about a special value as done in Section 6.2.  *Table 6.3* presents the names and descriptions of variables in this data set.
Finally, the third data set, `FLCensusDem`, contains the updated census data with two additional demographic variables -- gender and age. Unlike the other census data we analyzed earlier, each observation of this data set consists of one voting district and the proportion of each demographic by age, gender, and race within that district. The tables above present the names and descriptions of variables in this data set of Florida districts. There is also a table that contains the age groupings used in the variable names of the `FLCensusDem.csv` file.

## Question 1

Use Bayes' Rule to find a formula for the probability that a voter belongs to a given racial group conditional on their age, gender, surname, and residence location. Start by representing the quantity you're looking to calculate in probability notation, and then use Bayes' Rule to find the formula that could calculate this. 
Given the data sets available to us, can we use this formula to predict each voter's race?  If the answer is yes, briefly explain how you would make the prediction. If the answer is no, briefly explain why you cannot apply the formula you derived.

## Answer 1

$$
\frac{P(\text{age, gender, residence, surname} \mid \text{race}) 
      P(\text{race})}{P(\text{age, gender, residence, surname})} 
      =
  \frac{P(\text{age, gender, residence | race and surname})
    P(\text{race | surname})}{{P(\text{age, gender, residence | surname})}}
$$
Yes, with some manipulation of the data sets above, this formula can be used to predict each voter's race. If you calculate the probability of all age groups, gender, county, and surname, and then the probability of each race, you can use Bayes formula to predict each voter's race. The surname has to be taken independently as in the census data the surname has to be taken from a different dataset.

## Question 2

Assume that, given the person's race, the surname is conditionally independent from residence, age, and gender. Express this assumption in probability notation and also substantively interpret. Then, explain how under this assumption, the probability that a voter belongs to a given racial group conditional on their age and gender as well as their surname and residence location is given by the following formula.

$$
\frac{P(\text{residence}, \text{age}, \text{gender} \mid \text{race}) 
      P(\text{race} \mid \text{surname})}{P(\text{residence}, \text{age}, \text{gender} \mid \text{surname})}    
$$

## Answer 2

Surname is conditionally independent from residence, age, and gender
$$
P(\text{residence, age, gender| race and surname})
   =P(\text{residence, age, gender} \mid \text{race})
$$
Because the surname is conditionally independent from residence, age, and gender, it implies a joint probability surname given race, and the residence, age, and gender given race. The probability can be found separately and then multiplied.

This new assumption can be given by the new formula because the joint probability is the top of the fraction with the probability of residence, age, gender given surname.     



## Question 3

Using the formula derived in the previous question, we wish to compute the predicted probability that a voter belongs to a given racial group, conditional on their age and gender as well as their surname and residence location.  Provide a step-by-step explanation of how you would do this computation with the data available to us (don't actually use the data yet). 

Hint: you will need to modify the formula without invoking an additional assumption such that all quantities can be computed from the data sets we have. The definition of conditional probability and the law of total probability might be useful to break down parts of the equation into quantities we can calculate. 

## Answer 3
$$
\frac {
 \frac{P(\text{residence, age, gender, race}) }
 {P(\text{race})} P(\text{race | surname})}
 {\sum_{race}P(\text{residence}, \text{age}, \text{gender} \mid \text{race})P(\text{race | surname})}
$$
Race should be below the sigma in the equation above.

First calculate, P(residence, age, gender, race) as a proportion of the total population, and then divide by P(race). Next, calculate P(race | surname). Take those two quantities and multiply them together. This should be done for each race Then find the P(residence, age, gender | race) for each race and the P(race | surname) for each race and add those to quantities up together. You'll be left with 6 numbers of the numerator (one for each race), and a 1 number for the denominator. Each number for each race should be divided by the denominator creating 6 numbers that give you the probability of a voter's race given their residence, age, and gender or surname (as surname is independent from residence, age, and gender)


## Question 4

Use the procedure described in the previous question, compute the predicted probability for each voter in the `FLVoters.csv` that the voter belongs to a given racial group conditional on their age, gender, surname and residence location. Your final output should be a dataframe containing all observations with their predicted probabilities for each race. Exclude the voters with missing data from your analysis. Also, note that the csv file `cnames.csv` has been processed from `names.csv` using the code from Section 6.1. Thus, there is no need to re-adjust the values to account for negligibly small race percentages, but the racial proportions by surname are initially expressed as percentages rather than as decimals.

## Answer 4
```{r}
census <- read.csv("FLCensusDem.csv")

resrag.prop <- census[,4:183] * census$total.pop / sum(census$total.pop)

race.prop <- c(sum(rowSums(resrag.prop[,1:30])),
               sum(rowSums(resrag.prop[,31:60])),
               sum(rowSums(resrag.prop[,61:90])),
               sum(rowSums(resrag.prop[,91:120])),
               sum(rowSums(resrag.prop[,121:150])),
               sum(rowSums(resrag.prop[,151:180])))

names(race.prop) <- c("white", "black", "hispanic", "asian", "other", "mixed")

resag.rcond <- cbind(census[, 1:3],
                     resrag.prop[,1:30] / race.prop[1],
                     resrag.prop[,31:60] / race.prop[2],
                     resrag.prop[,61:90] / race.prop[3],
                     resrag.prop[,91:120] / race.prop[4],
                     resrag.prop[,121:150] / race.prop[5],
                     resrag.prop[,151:180] / race.prop[6])

cnames <- read.csv("cnames.csv")
cnames[,3:8] <- cnames[,3:8] / 100

voters <- read.csv("FLvoters.csv")

agebracket <- cbind(18:99, c(rep(1,2), rep(2,5), rep(3,5), rep(4,5), rep(5,5), 
                             rep(6,5), rep(7,5), rep(8,5), rep(9,5), rep(10,5), 
                             rep(11,5), rep(12,5),rep(13,5), rep(14,5), rep(15,5)))

head(agebracket)
colnames(agebracket) <- c("age", "agegrp")
voters <- merge (voters, agebracket, by = "age")
voters <- voters[is.na(voters$gender) == FALSE & is.na(voters$age) == FALSE,]

voters$category <- paste(voters$gender, voters$agegrp, sep = ".")
resag.rvoter <- matrix(NA, nrow=nrow(voters), ncol = 6)
colnames(resag.rvoter) <- c("resag.whi", "resag.bla", "resag.his", "resag.asi", 
                            "resag.oth", "resag.mix" )

voters <- cbind(voters, resag.rvoter)

for( i in c(1:nrow(voters))) {
                selected_variables <- c(paste(voters$category[i], "whi", sep = "."),
                         paste(voters$category[i], "bla", sep = "."),
                         paste(voters$category[i], "his", sep = "."),
                         paste(voters$category[i], "asi", sep = "."),
                         paste(voters$category[i], "oth", sep = "."),
                         paste(voters$category[i], "mix", sep = "."))
  temp <- subset(resag.rcond, subset = county == voters$county[i] & VTD == voters$VTD[i], 
                 select = selected_variables)
  voters[i, 9:14] <- temp
                              
}

voters2 <- merge(voters, cnames[, c(1, 3:4, 8, 5:7)], by = "surname")

resag.scond <- voters2[, 9:14] * voters2[, 15:20]
colnames(resag.scond) <- c("white", "black", "hispanic", "asian", "other", "mixed")

voters2 <- cbind(voters2, resag.scond)

head(voters2)


```


## Question 5

Given the predicted probabilities from the previous question, identify the most likely race for each individual in `FLVoters.csv`. Then, compare these predictions with the self-reported races of the voters, specifically for white, black, Hispanic, and Asian voters. Calculate the true positive rate and the false positive rate for each of these four racial groups. How does your model compare to the predictions made in Section 6.1 based on surname and residence location alone? 

## Answer 5
```{r}
voters2$predict <- rep(NA, nrow(temp))
voters2$max <- apply(voters2[,21:26], 1, max)

voters2$predict <- ifelse(voters2$max == voters2$white, "white", 
                       ifelse(voters2$max == voters2$black, "black",
                              ifelse(voters2$max == voters2$hispanic, "hispanic",
                                     ifelse(voters2$max == voters2$asian, "asian",
                                            ifelse(voters2$max == voters2$other, "other",
                                                    "mixed")))))

##true positives 
white <- voters2[voters2$race == "white",]
mean(voters2$predict == "white", na.rm = T)

black <- voters2[voters2$race == "black",]
mean(voters2$predict == "black", na.rm = T)

hispanic <- voters2[voters2$race == "hispanic",]
mean(voters2$predict == "hispanic", na.rm = T)

asian <- voters2[voters2$race == "asian",]
mean(voters2$predict == "asian", na.rm = T)




##false positives
predict.white <- voters2[voters2$predict == "white",]
mean(predict.white$race != "white", na.rm = T)

predict.black <- voters2[voters2$predict == "black",]
mean(predict.black$race != "black", na.rm = T)

predict.hispanic <- voters2[voters2$predict == "hispanic",]
mean(predict.hispanic$race != "hispanic", na.rm = T)

predict.asian <- voters2[voters2$predict == "asian",]
mean(predict.asian$race != "asian", na.rm = T)






```

